Trends from 30-year observations of downward solar irradiance in Thessaloniki, Greece

### 2023-12-19 Build_169/MDPI_submission_B169.tex
Commit 2023-12-19 11:09

The all-sky trend is similar to the one reported in \citet{Bais2013}
from a ten-year shorter dataset suggesting that the tendency of SDR in
Thessaloniki is systematic. Other studies for the European region
reported a change in the SDR trend around 1980 from negative to positive
with comparable magnitude \citep{Wild2021, Yuan2021, Ohmura2009}, well
before the start of our records. However, the trends reported here for
the three datasets are in accordance with the widely accepted solar
radiation brightening over Europe. For the period of our observations
the trend in the TSI is negligible (\(-0.00024\,\%/y\)), and thus we
cannot attribute any significant effect on the SDR trend to solar
variability.

@@

\caption{\label{tab:trendtable}Trends in SDR daily means for different sky conditions for the period 1993 - 2023.}
\begin{tabu} to \linewidth {>{\centering\arraybackslash}p{8em}>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X}
\toprule
Sky conditions & Trend [\%/year] & Trend S.E. ($2\sigma$) & Pearson cor. & Trend [W/year] & Days\\
\midrule
All skies & 0.3760 & 0.1210 & 0.0911 & 1.450 & 10250\\
Clear skies & 0.0972 & 0.0327 & 0.1380 & 0.501 & 2684\\
Cloudy skies & 0.4080 & 0.1770 & 0.0810 & 1.180 & 4937\\
\bottomrule
\end{tabu}
\end{table}

@@

Although the year-to-year variability of the anomalies (Figure
\ref{fig:trendALL} and Figures \ref{fig:trendCLEAR},
\ref{fig:trendCLOUD} in Appendix), shows a rather homogeneous behaviour,
plots of the cumulative sums (CUSUM) \citep{Regier2019} of the anomalies
can reveal different structures in the records of all three sky
conditions. For time series with a uniform trend, we would expect the
CUSUMs of the anomalies to have a symmetric `V' shape centered around
the middle of the series. This would indicate that the anomalies are
evenly distributed around the climatological mean, and for a positive
uniform trend, the first half is below and the second half above the
climatological mean. In our case, there is a more complex evolution of
the anomalies. For all-skies (Figure \ref{fig:cusummonth-1}) we observe
three rather distinct periods: a) a downward part between the start of
the datasets and about 2000, denoting that all anomalies are negative
thus below the climatology; b) a relatively steady part lasting for
almost 20 years suggesting little variability in SDR anomalies; and c) a
steep upward part to present indicating anomalies above the climatology.

(Figures \ref{fig:cusummonth-1} and \ref{fig:cusummonth-2}), we observe
three macroscopic periods. A downward part from the start until about
2005, a relatively steady part until about 2016 and, finally, a steep
upward part until the present. For cloud-sky
(Figure~\ref{fig:cusummonth-3}), we have a different pattern; it begins
with a relatively steady part until 1997, followed by an upward part
until 2005, and a long decline until 2020, with a small positive slope
until the present. For a uniform trend, we would expect the CUSUMs of
the anomalies to have a symmetric `V' shape. This would indicate that
the anomalies are evenly distributed around the climatological mean, and
for a positive uniform trend, the first half to be below and the other
half above the climatological mean. In our case, there is a more complex
evolution of the anomalies. Another distinct feature of the CUSUMs, is
the different pattern of the cloudy-sky dataset which peaks around the
middle of the period. Although, there seems to exist a complementary
relation to the CUSUMs of the clear- and all-sky cases, we can not
assert that clouds are the main driver for this relation due to the
great difference in the number of observational data between the two
datasets (Table~\ref{tab:trendtable}).

@@

In order to investigate further the features of the CUSUMs, we created
another set of CUSUM plots by subtracting the corresponding long term
trend from the SDR anomaly data, prior to the CUSUM calculation
(Figure~\ref{fig:cusumnotrendmonthly}). With this approach periods when
the CUSUMs diverge from zero can be interpreted as a systematic
variation of SDR from the climatological mean. When the CUSUM is
increasing, the added values are above the climatological values of the
SDR trend and vice versa. Overall, for all- and clear-sky conditions
(Figures~\ref{fig:cusumnotrendmonthly-1}
and~\ref{fig:cusumnotrendmonthly-2}) we observe periods when the
anomalies diverge from the climatological value, each lasting for
several years. The pattern in both datasets is very similar, suggesting
prevalence in clear skies over Thessaloniki. It is interesting that in
the period 1993 - 2016 the anomalies have a high variability around
zero, while after 2016, the range of the variability is decreased to
about one third of the prior period. For cloudy-sky conditions
(Figure~\ref{fig:cusumnotrendmonthly-3}) the period 1997 - 2008 is
dominated by positive CUSUMs, suggesting a reduced effect of clouds on
SDR. From 1997 to mid-2000s CUSUMs are increasing, likely due to a
continuous decrease in the optical thickness of clouds, followed by a
period of rapid increase (within 3 years) in cloud optical thickness
lasting up to 2008. The following stable period spans for about 15 years
up to 2021 when CUSUMs start increasing again.

@@

Similarly to the long term trends from daily means of SDR discussed
above, we have calculated the trend for the three sky conditions and for
each season of the year, using the corresponding mean monthly anomalies
(Figure~\ref{fig:seasonalALL} and Table~\ref{tab:trendseasontable}). For
all-sky conditions the trend in SDR in winter is the largest
(\(0.7\,\%/y\)), followed by the trend in autumn (\(0.42\,\%/y\), a
value close to the long term trend) both statistically significant at
the \(99\,\%\) confidence level. In spring and summer, the trends are
much smaller and of lesser statistical significance. These seasonal
differences indicate a possible relation of the trends in SDR to trends
of clouds during winter and autumn. For clear-skies, the trend in winter
is \(0.4\,\%/y\), larger than for all-skies (\(0.7\,\%/y\)), which is
another indication of a decreasing trend in cloud optical thickness.
Moreover, the trends under clear- and cloudy-sky conditions are almost
complementary to each other, particularly for winter and autumn, where
the signal is stronger. During spring and summer the statistical
significance is very low and the actual trend too small for a meaningful
comparison.

@@

\caption{\label{tab:trendseasontable}SDR linear trends of monthly anomalies for each season of the year.}
\begin{tabu} to \linewidth {>{\centering\arraybackslash}p{8em}>{\centering}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X}
\toprule
Sky condition & Season & Trend [\%/year] & Pearson cor. & Trend p-value\\
\midrule
\cellcolor{gray!6}{} & \cellcolor{gray!6}{Winter} & \cellcolor{gray!6}{0.7000} & \cellcolor{gray!6}{0.5350} & \cellcolor{gray!6}{0.002770}\\

 & Spring & 0.1070 & 0.1660 & 0.371000\\

\cellcolor{gray!6}{} & \cellcolor{gray!6}{Summer} & \cellcolor{gray!6}{0.1050} & \cellcolor{gray!6}{0.2540} & \cellcolor{gray!6}{0.175000}\\

\multirow{-4}{*}{\centering\arraybackslash All skies} & Autumn & 0.4230 & 0.4660 & 0.009390\\
\cmidrule{1-5}
\cellcolor{gray!6}{} & \cellcolor{gray!6}{Winter} & \cellcolor{gray!6}{0.3950} & \cellcolor{gray!6}{0.5960} & \cellcolor{gray!6}{0.000515}\\

 & Spring & 0.0591 & 0.1270 & 0.497000\\

\cellcolor{gray!6}{} & \cellcolor{gray!6}{Summer} & \cellcolor{gray!6}{-0.0540} & \cellcolor{gray!6}{-0.3010} & \cellcolor{gray!6}{0.106000}\\

\multirow{-4}{*}{\centering\arraybackslash Clear skies} & Autumn & 0.0539 & 0.1710 & 0.366000\\
\cmidrule{1-5}
\cellcolor{gray!6}{} & \cellcolor{gray!6}{Winter} & \cellcolor{gray!6}{0.7590} & \cellcolor{gray!6}{0.5860} & \cellcolor{gray!6}{0.000665}\\

 & Spring & 0.0621 & 0.0998 & 0.593000\\

\cellcolor{gray!6}{} & \cellcolor{gray!6}{Summer} & \cellcolor{gray!6}{-0.0789} & \cellcolor{gray!6}{-0.1110} & \cellcolor{gray!6}{0.560000}\\

\multirow{-4}{*}{\centering\arraybackslash Cloudy skies} & Autumn & 0.1910 & 0.1650 & 0.384000\\
\bottomrule
\end{tabu}
\end{table}

@@

We have analyzed a 30 year dataset of SRD measurements in Thessaloniki
Greece (1993 -- 2023) aiming to identify the long term variability of
solar irradiance under different sky conditions. Under all-sky
conditions there is a positive trend in SDR of \(0.38\,\%/y\)
(brightening). A previous study \citep{Bais2013} for the period 1993 --
2011 found also a positive trend of \(0.33\,\%/y\). The increase of this
trend indicates that the brightening of SDR continues and is probably
caused by continuing decreases in aerosol optical depth and the optical
thickness of clouds over the area. Moreover, we found a similar trend
under clear-sky conditions (\(0.1\,\%/y\)) that further supports the
assumption that the brightening is caused mainly by decreasing aerosols.
Unfortunately, for the entire period there is no available data for the
aerosols, in order to quantify their effect on SDR. However,
\citet{Siomos2020} have shown that aerosol optical depth over
Thessaloniki is decreasing constantly at least up to 2018. The
attenuation of SDR by aerosols over Europe have been proposed as major
factor by \citet{Wild2021}. The dimming effect on SDR under cloudy-sky
conditions (\(0.41\,\%/y\)), suggests that cloud optical thickness is
decreasing during this period. Because we have no adequate data to
investigate the long term changes of cloud thickness in the region, we
cannot verify if the negative SDR trend we observe under under
cloudy-skies can be attributed solely to changes in clouds.

>>>>

The all-sky trend is similar to the one reported in \citet{Bais2013}
from a ten-year shorter dataset suggesting that the tendency of SDR in
Thessaloniki is systematic. Other studies for the European region
reported a change in the SDR trend around 1980 from negative to positive
with comparable magnitude \citep{Wild2021, Yuan2021, Ohmura2009}, well
before the start of our records. However, the trends reported here for
the three datasets are in accordance with the widely accepted solar
radiation brightening over Europe. For the period of our observations
the trend in the TSI is negligible (\(-0.00022\,\%/y\)), and thus we
cannot attribute any significant effect on the SDR trend to solar
variability.

@@

\caption{\label{tab:trendtable}Trends in SDR daily means for different sky conditions for the period 1993 - 2023.}
\begin{tabu} to \linewidth {>{\centering\arraybackslash}p{8em}>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X}
\toprule
Sky conditions & Trend [\%/year] & Trend S.E. ($2\sigma$) & Pearson correl. & Trend [W/m\textsuperscript{2}/year] & Days\\
\midrule
All skies & 0.3760 & 0.1210 & 0.0913 & 1.460 & 10251\\
Clear skies & 0.0972 & 0.0327 & 0.1380 & 0.501 & 2684\\
Cloudy skies & 0.4080 & 0.1770 & 0.0810 & 1.180 & 4937\\
\bottomrule
\end{tabu}
\end{table}

@@

Although the year-to-year variability of the anomalies (Figure
\ref{fig:trendALL} and Figures~\ref{fig:trendCLEAR},
\ref{fig:trendCLOUD} in Appendix), shows a rather homogeneous behaviour,
plots of the cumulative sums (CUSUM) \citep{Regier2019} of the anomalies
can reveal different structures in the records of all three sky
conditions. For time series with a uniform trend, we would expect the
CUSUMs of the anomalies to have a symmetric `V' shape centered around
the middle of the series. This would indicate that the anomalies are
evenly distributed around the climatological mean, and for a positive
uniform trend, the first half is below and the second half above the
climatological mean. In our case, there is a more complex evolution of
the anomalies. For all-skies (Figure~\ref{fig:cusummonth-1}), we observe
three rather distinct periods: a) a downward part between the start of
the datasets and about 2000, denoting that all anomalies are negative
thus below the climatology; b) a relatively steady part lasting for
almost 20 years suggesting little variability in SDR anomalies; and c) a
steep upward part to present indicating anomalies above the climatology.
The CUSUMs for cloudy-skies (Figure~\ref{fig:cusummonth-3}), show a
similar behavior with some short-term differences that do not change the
overall pattern. For clear skies (Figure~\ref{fig:cusummonth-2}), a
monotonic downward tendency is evident until 2004, suggesting that the
anomalies are all negative. After 2004 the anomalies turn to positive at
a fast rate for about five years and at a slower rate thereafter.

@@

In order to unveil further the features of the variability of the three
datasets, Figure~\ref{fig:cusumnotrendmonthly} of the CUSUMs, we created
another set of CUSUM plots by subtracting the corresponding long term
trend from the SDR anomaly data, prior to the CUSUM calculation
(Figure~\ref{fig:cusumnotrendmonthly}). With this approach periods when
the CUSUMs diverge from zero can be interpreted as a systematic
variation of SDR from the climatological mean. When the CUSUM is
increasing, the added values are above the climatological values of the
SDR trend and vice versa. Overall, for all- and clear-sky conditions
(Figures~\ref{fig:cusumnotrendmonthly-1}
and~\ref{fig:cusumnotrendmonthly-2}) we observe periods when the
anomalies diverge from the climatological value, each lasting for
several years. The pattern in both datasets is very similar, suggesting
prevalence in clear skies over Thessaloniki. It is interesting that in
the period 1993 - 2016 the anomalies have a high variability around
zero, while after 2016, the range of the variability is decreased to
about one third of the prior period. For cloudy-sky conditions
(Figure~\ref{fig:cusumnotrendmonthly-3}) the period 1997 - 2008 is
dominated by positive CUSUMs, suggesting a reduced effect of clouds on
SDR. From 1997 to mid-2000s CUSUMs are increasing, likely due to a
continuous decrease in the optical thickness of clouds, followed by a
period of rapid increase (within 3 years) in cloud optical thickness
lasting up to 2008. The following stable period spans for about 15 years
up to 2021 when CUSUMs start increasing again.

@@

Similarly to the long term trends from daily means of SDR discussed
above, we have calculated the trend for the three sky conditions and for
each season of the year, using the corresponding mean monthly anomalies
(Figure~\ref{fig:seasonalALL} and Table~\ref{tab:trendseasontable}). For
all-sky conditions the trend in SDR in winter is the largest
(\(0.7\,\%/y\)), followed by the trend in autumn (\(0.42\,\%/y\), a
value close to the long term trend) both statistically significant at
the \(99\,\%\) confidence level. In spring and summer, the trends are
much smaller and of lesser statistical significance. These seasonal
differences indicate a possible relation of the trends in SDR to trends
in clouds during winter and autumn. For clear-skies, the trend in winter
is \(0.4\,\%/y\), larger than for all-skies (\(0.7\,\%/y\)), which is
another indication of a decreasing trend in cloud optical thickness.
Moreover, the trends under clear- and cloudy-sky conditions are almost
complementary to each other, particularly for winter and autumn, where
the signal is stronger. During spring and summer the statistical
significance is very low and the actual trend too small for a meaningful
comparison.

@@

\caption{\label{tab:trendseasontable}SDR linear trends of monthly anomalies for each season of the year.}
\begin{tabu} to \linewidth {>{\centering\arraybackslash}p{8em}>{\centering}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X}
\toprule
Sky condition & Season & Trend [\%/year] & Trend S.E. ($2\sigma$) & Pearson correl. & Trend p-value\\
\midrule
\cellcolor{gray!6}{} & \cellcolor{gray!6}{Winter} & \cellcolor{gray!6}{0.7000} & \cellcolor{gray!6}{0.4250} & \cellcolor{gray!6}{0.5350} & \cellcolor{gray!6}{0.002770}\\

 & Spring & 0.1070 & 0.2350 & 0.1660 & 0.371000\\

\cellcolor{gray!6}{} & \cellcolor{gray!6}{Summer} & \cellcolor{gray!6}{0.1050} & \cellcolor{gray!6}{0.1520} & \cellcolor{gray!6}{0.2540} & \cellcolor{gray!6}{0.175000}\\

\multirow{-4}{*}{\centering\arraybackslash All skies} & Autumn & 0.4230 & 0.3030 & 0.4660 & 0.009390\\
\cmidrule{1-6}
\cellcolor{gray!6}{} & \cellcolor{gray!6}{Winter} & \cellcolor{gray!6}{0.3950} & \cellcolor{gray!6}{0.2010} & \cellcolor{gray!6}{0.5960} & \cellcolor{gray!6}{0.000515}\\

 & Spring & 0.0591 & 0.1720 & 0.1270 & 0.497000\\

\cellcolor{gray!6}{} & \cellcolor{gray!6}{Summer} & \cellcolor{gray!6}{-0.0540} & \cellcolor{gray!6}{0.0646} & \cellcolor{gray!6}{-0.3010} & \cellcolor{gray!6}{0.106000}\\

\multirow{-4}{*}{\centering\arraybackslash Clear skies} & Autumn & 0.0539 & 0.1170 & 0.1710 & 0.366000\\
\cmidrule{1-6}
\cellcolor{gray!6}{} & \cellcolor{gray!6}{Winter} & \cellcolor{gray!6}{0.7590} & \cellcolor{gray!6}{0.3970} & \cellcolor{gray!6}{0.5860} & \cellcolor{gray!6}{0.000665}\\

 & Spring & 0.0621 & 0.2300 & 0.0998 & 0.593000\\

\cellcolor{gray!6}{} & \cellcolor{gray!6}{Summer} & \cellcolor{gray!6}{-0.0789} & \cellcolor{gray!6}{0.2670} & \cellcolor{gray!6}{-0.1110} & \cellcolor{gray!6}{0.560000}\\

\multirow{-4}{*}{\centering\arraybackslash Cloudy skies} & Autumn & 0.1910 & 0.4320 & 0.1650 & 0.384000\\
\bottomrule
\end{tabu}
\end{table}

@@

We have analyzed a 30 year dataset of SRD measurements in Thessaloniki
Greece (1993 -- 2023) aiming to identify the long term variability of
solar irradiance under different sky conditions. Under all-sky
conditions there is a positive trend in SDR of \(0.38\,\%/y\)
(brightening). A previous study \citep{Bais2013} for the period 1993 --
2011 reported also a positive trend, of \(0.33\,\%/y\). The slight
increase of this trend indicates that the brightening of SDR continues
and is probably caused by continuing decreases in aerosol optical depth
and the optical thickness of clouds over the area. A smaller trend has
been found under clear-sky conditions (\(0.1\,\%/y\)) that further
supports the assumption that the brightening is caused mainly by
decreasing aerosols. Unfortunately, for the entire period there is no
available data for the aerosols, in order to quantify their effect on
SDR. However, \citet{Siomos2020} have shown that aerosol optical depth
over Thessaloniki is decreasing constantly at least up to 2018. The
attenuation of SDR by aerosols over Europe have been proposed as major
factor by \citet{Wild2021}. The dimming effect on SDR under cloudy-sky
conditions (\(0.41\,\%/y\)), suggests that cloud optical thickness is
decreasing during this period. Because we have no adequate data to
investigate the long term changes of cloud thickness in the region, we
cannot verify if the negative SDR trend we observe under under
cloudy-skies can be attributed solely to changes in clouds.
